HO unification from object language to meta language

### 2024-04-03 paper/main.tex
tex: update beta section

\begin{lemma}
  \elpiIn{progress} terminates.
\end{lemma}

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape $X~t_1\dots t_n$ such that $X$ is a
  flexible variable and $t_1\dots t_n$ is not in \llambda.
  \todo[inline]{si dice: flexible term; oppure unification variable}
\end{invariant}

@@

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta,
  the current unification problem is not in \llambda
  \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
  and the unification fails.
\end{corollary}

@@

The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having in scope all the free
variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
scope all the free variables occurring in \elpiIn{Pf1}
\todo[inline]{anche in Extra1 no?}.

\newcommand{\rhsBeta}{\ensuremath{N_S\appsep L}\xspace}
\todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
... where S = [s1 .. sn] e L = [l1 .. lm]}

\begin{invariant}
  The \rhs of a \linkbeta has the shape \rhsBeta (equivalent to
  \elpiIn{app[uva N S|L]}).
\end{invariant}

\begin{corollary}
  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then \elpiIn{L} is not
  empty.
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, if \elpiIn{L} is a empty list, then the original \Fo{} term
  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
  distinct names (i.e. there list \elpiIn{Extra} is empty). This case is however
  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
  \linkbeta is produced: a contradiction.
\end{proof}

\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then the 
  first argument $t$ in \elpiIn{L} either appears in \elpiIn{S} or it is 
  not a name.
\end{corollary}

\begin{proof}[Proof sketch]
  By construction, the lists \elpiIn{S} and \elpiIn{L} are built by splitting
  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
  \elpiIn{S} is the longest prefix of the compiled terms in \elpiIn{Ag} which is
  in \llambda. Therefore, by definition of \llambda, either the first element of
  \elpiIn{L} is a name appearing in \elpiIn{S} or it a term with a constructor
  of \elpiIn{tm} as functor.  
\end{proof}

@@

\begin{definition}[\progBetaLL]
  A link \linkbetaM{\Gamma}{X}{N_S\appsep [t|L]} is removed from \linkStore, if
  it exists a substitution $\gamma$, such that the first term $t$ is a name not
  occurring in $S$. In this case, let $M$ a fresh variable, then the new link
  \linkbetaM{\Gamma,x}{X}{M_{S,n}\appsep L} is added to \linkStore and if $L =
  []$, then $X$ is unified with $M_{S,n}$, otherwise, the
  \linketaM{\Gamma}{N}{\lambda x.M_x} is added to \linkStore.
  \label{def:progBetaLL}
  \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
\end{definition}

\begin{definition}[\progBetaRH]
  A link \linkbetaM{\Gamma}{X}{\rhsBeta} is removed from
  \linkStore if $N_S$ is instantiated to a term $t$ and the
  $\beta$-reduced term $t'$ obtained from the application of $t$ to
  $L$ is in \llambda. Moreover, $X$ is unified to $t$.
  \label{def:progBetaRH}
\end{definition}

@@

\begin{proof}[Proof sketch]
  By \cref{def:progBetaRH}, the \linkbeta is removed from \linkStore, hence 
  they cannot be applied indefinitely. If the \linkbeta is progressed due to
  \progBetaLL, then two links are added, the first
  \todo[inline]{finire}
\end{proof}

\begin{corollary}
  Given a \linkbeta, the scope of its \rhs variables in \llambda.
  \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
\end{corollary}

\begin{proof}[Proof sketch]
  By construction, the \rhs of \linkbeta is of the form \rhsBeta and $S$ is in
  \llambda. If a \linkbeta is triggered by \progBetaRH, then, by
  \cref{def:progBetaRH} that link is removed by \linkStore, and the property is
  satisfied. If the \linketa is activated by \progBetaLL, then, by
  \cref{def:progBetaLL}, $L = [n|L']$ and the scope `$S,x$', is in \llambda. 
\end{proof}

>>>>

\begin{lemma}
  \elpiIn{progress} terminates.
  \label{lemma:prog-eta-terminates}
\end{lemma}

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape $X~t_1\dots t_n$ such that $X$ is a
  flexible term and $t_1\dots t_n$ is not in \llambda.
\end{invariant}

@@

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta,
  the original unification problem is not in \llambda
  % \todo[inline]{non capisco cosa vuoi dire con "the current...". Data la prova sotto vuoi dire che il problem originale sarebbe stato fuori... e quindi falliamo}
  \todo[inline]{Is it clearer?}
  and the unification fails.
\end{corollary}

@@

The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having in scope all the free
variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}

\newcommand{\rhsBetaHead}{\ensuremath{X_{s_1\dots s_n}}}
\newcommand{\rhsBeta}{\ensuremath{\rhsBetaHead\appsep t_1\dots t_m}\xspace}
% \todo[inline]{Io direi $X_{s_1\ldots s_n}\appsep l_1\ldots l_m$
% ... where S = [s1 .. sn] e L = [l1 .. lm]}
\todo[inline]{Is it clearer?}

In the following, we pose $X_{s_1\dots s_n}$ to represent a variable $X$ with
the list of distinct names $s_1\dots s_n$ as scope. Moreover, $X_{s_1\dots
s_n}\appsep t_1\dots t_m$ is the application of that variable to the list of
terms $t_1\dots t_m$. Note that \rhsBeta is equivalent to \elpiIn{app[uva N S |
T]}.

\begin{invariant}
  The \rhs of a \linkbeta has the shape \rhsBeta.
\end{invariant}

\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, if $m = 0$, then the original \Fo{} term
  has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of
  distinct names (i.e. the list \elpiIn{Extra} is empty). This case is however
  captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and no
  \linkbeta is produced: a contradiction.
\end{proof}

\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
  not a name.
\end{corollary}

\begin{proof}[Proof sketch]
  By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
  $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
  in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
  in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
  \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
\end{proof}

@@

\begin{definition}[\progBetaLL]
  A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  the substitution $\sigma$, $\sigma l_1$ is a name, say $t$, such that, $t
  \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % \todo[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  \todo[inline]{Is it clearer?}
\end{definition}

\begin{definition}[\progBetaRH]
  A link \linkbetaM{\Gamma}{X}{\rhsBeta} is removed from
  \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
  $\beta$-reduced term $t'$ obtained from the application of $t$ to
  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified to $t$.
  \label{def:progBetaRH}
\end{definition}

@@

\begin{proof}[Proof sketch]
  \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  creates each time a new \linketa and replace the old \linkbeta with a new one.
  This progression can however triggered at most $m$ times, since each new
  \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  the \linkbeta is removed from \linkStore which is now filled by $m$ new
  \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  terminates if \linkStore is made by only \linketa and therefore
  \elpiIn{progress} still terminates.
  \todo[inline]{is it ok?}
\end{proof}

\begin{corollary}
  Given a \linkbeta, the variables occurring in its \rhs is in \llambda.
  % \todo[inline]{vuoi dire: le variabili che occorrono nel rhs sono in...?}
  \todo[inline]{is it clearer?}
\end{corollary}

\begin{proof}[Proof sketch]
  By construction, the \rhs of \linkbeta is of the form \rhsBeta, $s_1\dots s_n$
  is in \llambda and all the terms $t_1\dots t_n$ are in \llambda, too. If a
  \linkbeta is triggered by \progBetaRH, then, by \cref{def:progBetaRH}, that
  link is removed by \linkStore, and the property is satisfied. If the \linketa
  is activated by \progBetaLL, then, by \cref{def:progBetaLL}, the new \linkbeta
  as a variable as a scope which is still in \llambda. 
\end{proof}
